---
title: "Code"
author: "Emily Mittleman & Julia Rosner"
date: '2022-12-05'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(broom)
library(stringr)
library(knitr)
library(nnet)
library(ggplot2)
library(MASS)
library(ISLR)
library(leaps)
library(glmnet)
library(mgcv)
library(car)
library(splines)
library(ggpubr)
library(corrplot)
library(rpart)
library(caret)
library(randomForest)
```


This file will be used for our initial code while we explore the data, different models, etc. Then we'll compile it into Report.Rmd

### Load Data
```{r load data}
data <- read.csv("Data/diabetes_binary_5050split.csv", header = TRUE)
colnames(data)[colnames(data) == "Diabetes_binary"] = "diabetes"

data.large <- read.csv("Data/diabetes_binary.csv", header = TRUE)
colnames(data.large)[colnames(data.large) == "Diabetes_binary"] = "diabetes"
```

### EDA

Look at correlations between variables. helps to know which attributes are highy dependent on the prediction variable
```{r correlations}
correlations <- cor(data)
corrplot(correlations, method="color")
```

Next, look at box plots of the 2 most correlated predictors and color by outcome.
```{r}
ggplot(data, aes(x = HighBP, fill = factor(diabetes))) +
  geom_bar(position="fill")
```
```{r}
ggplot(data, aes(x = GenHlth, fill = factor(diabetes))) +
  geom_bar(position="fill")
```
```{r}
ggplot(data, aes(x = GenHlth, fill = factor(HighBP))) +
  geom_bar(position="fill")
```
Make pivot table to make historgrams of each variable simpler
```{r}
data_long <- data %>%                          # Apply pivot_longer function
  pivot_longer(colnames(data)) %>% 
  as.data.frame()
head(data_long)
```
Visualize predictor variable distrbutions:

```{r fig.height=10, fig.width=15, fig.fullwidth=TRUE}
ggp1 <- ggplot(data_long, aes(x = value)) +    # Draw each column as histogram
  geom_histogram(bins=10) + 
  facet_wrap(~ name, scales = "free")+ 
  theme(text=element_text(size=20))
ggp1
```

Next, look for outliers in predictors:

```{r make boxplots}
diabetes_labels <- c('no', 'yes')

p1 <- ggplot(data, aes(x = BMI, y=factor(diabetes), color=factor(diabetes))) +
  geom_boxplot(outlier.shape=8, outlier.size=4)+
  labs(title="BMI", y="Diabetes")+
  scale_color_discrete(name="diabetes", labels=diabetes_labels)
p2 <- ggplot(data, aes(x = GenHlth, y=factor(diabetes), color=factor(diabetes))) +
  geom_boxplot(outlier.shape=8, outlier.size=4)+
  labs(title="GenHlth",y="Diabetes")+
  scale_color_discrete(name="diabetes", labels=diabetes_labels)
p3 <- ggplot(data, aes(x = MentHlth, y=factor(diabetes), color=factor(diabetes))) +
  geom_boxplot(outlier.shape=8, outlier.size=4)+
  labs(title="MentHlth", y="Diabetes")+
  scale_color_discrete(name="diabetes", labels=diabetes_labels)
p4 <- ggplot(data, aes(x = PhysHlth, y=factor(diabetes), color=factor(diabetes))) +
  geom_boxplot(outlier.shape=8, outlier.size=4)+
  labs(title="PhysHlth",y="Diabetes")+
  scale_color_discrete(name="diabetes", labels=diabetes_labels)
p5 <- ggplot(data, aes(x = Age, y=factor(diabetes), color=factor(diabetes))) +
  geom_boxplot(outlier.shape=8, outlier.size=4)+
  labs(title="Age",y="Diabetes")+
  scale_color_discrete(name="diabetes", labels=diabetes_labels)
p6 <- ggplot(data, aes(x = Education, y=factor(diabetes), color=factor(diabetes))) +
  geom_boxplot(outlier.shape=8, outlier.size=4)+
  labs(title="Education", y="Diabetes")+
  scale_color_discrete(name="diabetes", labels=diabetes_labels)
p7 <- ggplot(data, aes(x = Income, y=factor(diabetes), color=factor(diabetes))) +
  geom_boxplot(outlier.shape=8, outlier.size=4)+
  labs(title="Income", y="Diabetes")+
  scale_color_discrete(name="diabetes", labels=diabetes_labels)
p1
p2
p3
p4
p5
p6
p7
```
From the boxplots above, we see that the predictors BMI, MntHlth, and PhysHlth have a lot of outliers. All three distributions aee very skewed to the right. GenHlth and Age have only a couple outliers. Education and Income have none.

Now, we visulaize predictor distributions and relation to response.
```{r}
ggplot(data, aes(x = GenHlth, fill = factor(diabetes))) +
  geom_bar(position="fill")
ggplot(data, aes(x = GenHlth, fill = factor(diabetes))) +
  geom_bar(position="fill")
```

```{r stacked boxplot per variable}
pbox1 <- ggplot(data, aes(x = HighBP, fill=factor(diabetes))) +
  geom_bar(position="fill")+
  labs(title="HighBP", y="Diabetes")+
  scale_fill_discrete(name="diabetes", labels=diabetes_labels)
pbox2 <- ggplot(data, aes(x = HighChol, fill=factor(diabetes))) +
  geom_bar(position="fill")+
  labs(title="HighChol", y="Diabetes")+
  scale_fill_discrete(name="diabetes", labels=diabetes_labels)
pbox3 <- ggplot(data, aes(x = CholCheck, fill=factor(diabetes))) +
  geom_bar(position="fill")+
  labs(title="CholCheck", y="Diabetes")+
  scale_fill_discrete(name="diabetes", labels=diabetes_labels)
pbox4 <- ggplot(data, aes(x = Smoker, fill=factor(diabetes))) +
  geom_bar(position="fill")+
  labs(title="Smoker", y="Diabetes")+
  scale_fill_discrete(name="diabetes", labels=diabetes_labels)
pbox5 <- ggplot(data, aes(x = Stroke, fill=factor(diabetes))) +
  geom_bar(position="fill")+
  labs(title="Stroke", y="Diabetes")+
  scale_fill_discrete(name="diabetes", labels=diabetes_labels)
pbox6 <- ggplot(data, aes(x = HeartDiseaseorAttack, fill=factor(diabetes))) +
  geom_bar(position="fill")+
  labs(title="HeartDiseaseorAttack", y="Diabetes")+
  scale_fill_discrete(name="diabetes", labels=diabetes_labels)
pbox7 <- ggplot(data, aes(x = PhysActivity, fill=factor(diabetes))) +
  geom_bar(position="fill")+
  labs(title="PhysActivity", y="Diabetes")+
  scale_fill_discrete(name="diabetes", labels=diabetes_labels)
pbox8 <- ggplot(data, aes(x = Veggies, fill=factor(diabetes))) +
  geom_bar(position="fill")+
  labs(title="Veggies", y="Diabetes")+
  scale_fill_discrete(name="diabetes", labels=diabetes_labels)
pbox9 <- ggplot(data, aes(x = HvyAlcoholConsump, fill=factor(diabetes))) +
  geom_bar(position="fill")+
  labs(title="HvyAlcoholConsump", y="Diabetes")+
  scale_fill_discrete(name="diabetes", labels=diabetes_labels)
pbox10 <- ggplot(data, aes(x = AnyHealthcare, fill=factor(diabetes))) +
  geom_bar(position="fill")+
  labs(title="AnyHealthcare", y="Diabetes")+
  scale_fill_discrete(name="diabetes", labels=diabetes_labels)
pbox11 <- ggplot(data, aes(x = NoDocbcCost, fill=factor(diabetes))) +
  geom_bar(position="fill")+
  labs(title="NoDocbcCost", y="Diabetes")+
  scale_fill_discrete(name="diabetes", labels=diabetes_labels)
pbox12 <- ggplot(data, aes(x = DiffWalk, fill=factor(diabetes))) +
  geom_bar(position="fill")+
  labs(title="DiffWalk", y="Diabetes")+
  scale_fill_discrete(name="diabetes", labels=diabetes_labels)
pbox1
pbox2
pbox3
pbox4
pbox5
pbox6
pbox7
pbox8
pbox9
pbox10
pbox11
pbox12

```

Next, look at "Age" and its relation to response (diabetes diagnosis):

```{r}
ggplot(data, aes(x = Age, fill=factor(diabetes))) +
  geom_bar(position="dodge")+
  labs(title="Age")+
  scale_fill_discrete(name="diabetes", labels=diabetes_labels)
```


### EDA of data binary outcome dataset

Next, look at plots of 2 most correlated predictors and color by outcome.
```{r ggplot}
ggplot(data, aes(x = HighBP, fill = diabetes)) +
  geom_bar(position="fill")
```

```{r}
ggplot(data, aes(x = GenHlth, fill = diabetes)) +
  geom_bar(position="fill")
```

```{r}
ggplot(data, aes(x = GenHlth, fill = HighBP)) +
  geom_bar(position="fill")
```

## Modeling

### Data preparation

```{r}
# Factor variables
data$diabetes <- as.factor(data$diabetes)
data$HighBP <- as.factor(data$HighBP)
data$CholCheck <- as.factor(data$CholCheck)
data$Smoker <- as.factor(data$Smoker)
data$Stroke <- as.factor(data$Stroke)
data$HeartDiseaseorAttack <- as.factor(data$HeartDiseaseorAttack)
data$PhysActivity <- as.factor(data$PhysActivity)
data$Fruits <- as.factor(data$Fruits)
data$Veggies <- as.factor(data$Veggies)
data$HvyAlcoholConsump <- as.factor(data$HvyAlcoholConsump)
data$AnyHealthcare <- as.factor(data$AnyHealthcare)
data$NoDocbcCost <- as.factor(data$NoDocbcCost)
data$GenHlth <- as.factor(data$GenHlth)
data$MentHlth <- as.factor(data$MentHlth)
data$DiffWalk <- as.factor(data$DiffWalk)
data$Sex <- as.factor(data$Sex)
data$Age <- as.factor(data$Age)
data$Education <- as.factor(data$Education)
data$Income <- as.factor(data$Income)
```

```{r split data}
# Split data train and test:
set.seed(17)
sample <- sample(c(TRUE, FALSE), nrow(data), replace=TRUE, prob=c(0.7,0.3))
train  <- data[sample, ]
test   <- data[!sample, ]
```


### Logistic Regression

```{r loog reg}
glm.fit.all <- glm(diabetes ~ HighBP+ HighChol + CholCheck + HeartDiseaseorAttack + AnyHealthcare
+ PhysActivity + HvyAlcoholConsump + Fruits + Veggies + GenHlth + DiffWalk + Sex + Income + Education + BMI + PhysHlth, 
               data = data, family = binomial)
summary(glm.fit.all)
```
```{r}
glm.probs.all <- predict(glm.fit.all, type = "response")
glm.probs.all[1:10]
```


```{r}
glm.pred.all <- rep(0, length(data$diabetes))
glm.pred.all[glm.probs.all > 0.5] <- 1
```

```{r}
table(glm.pred.all, data$diabetes)
```

```{r}
accuracy <- sum(diag(table(glm.pred.all, data$diabetes)))/nrow(data)
accuracy
```

Now make model based off of training data:

```{r}
glm.fit.trainall <- glm(diabetes ~ HighBP+ HighChol + CholCheck + HeartDiseaseorAttack + AnyHealthcare
+ PhysActivity + HvyAlcoholConsump + Fruits + Veggies + GenHlth + DiffWalk + Sex + Income + Education + BMI + PhysHlth, 
               data = train, family = binomial)
glm.probs.trainall <- predict(glm.fit.trainall, test, type = "response")
```

```{r}
glm.pred.trainall <- rep(0, length(test))
glm.pred.trainall[glm.probs.trainall > 0.5] <- 1
table(glm.pred.trainall, test$diabetes)
```
```{r}
accuracy <- sum(diag(table(glm.pred.trainall, test$diabetes)))/nrow(test)
accuracy
```

To improve the accuracy we will consider a subset of predictors. Look at correlations to decide. The most correlated to diabetes are GenHlth and HighBP.

```{r}
glm.fit.cor <- glm(diabetes ~ GenHlth + HighBP, data=train, family = binomial)
glm.probs.cor <- predict(glm.fit.cor, test, type = "response")
glm.pred.cor <- rep("no diabetes", length(test))
glm.pred.cor[glm.probs.cor > 0.5] <- "diabetes"
table(glm.pred.cor, test$diabetes)
```
```{r}
Accuracy <- (0+5)/(1+18+5+0)
Accuracy
```

The subset of predictors made our predictive performance worse.

### KNN

```{r knn}
#KNN wont knit but works (just takes a while to run)
#library(class)
#set.seed(1)
#knn.pred <- knn(train, test, train$diabetes, k = 10)
#table(knn.pred, test$diabetes)
```

```{r}
#accuracy <- sum(diag(table(knn.pred, test$diabetes)))/nrow(test)
#accuracy
```

Perform CV to find best k value....?

### Trees

```{r}
library(tree)
```

```{r}
tree.all <- tree(diabetes ~ HighBP+ HighChol + CholCheck + HeartDiseaseorAttack + AnyHealthcare
+ PhysActivity + HvyAlcoholConsump + Fruits + Veggies + GenHlth + DiffWalk + Sex + Income + Education + BMI + PhysHlth, data)
summary(tree.all)
```

```{r}
plot(tree.all)
text(tree.all, pretty = 0)
```
```{r}
set.seed(3)
cv.tree.all <- cv.tree(tree.all, FUN = prune.misclass)
names(cv.tree.all)
```

```{r}
cv.tree.all
```
```{r}
par(mfrow = c(1,2))
plot(cv.tree.all$size, cv.tree.all$dev, type = "b")
plot(cv.tree.all$k, cv.tree.all$dev, type = "b")
```
```{r}
prune.tree <- prune.misclass(tree.all, best = 4)
plot(prune.tree)
text(prune.tree, pretty = 0)
```


### Decision Tree

```{r}
accuracy_tune <- function(fit) {
    predict_unseen <- predict(fit, test, type = 'class')
    table_mat <- table(test$diabetes, predict_unseen)
    accuracy_Test <- sum(diag(table_mat)) / sum(table_mat)
    accuracy_Test
}

control <- rpart.control(maxdepth = 15)
tune_fit <- rpart(diabetes ~ ., data = train, method = 'class', control = control)
accuracy_tune(tune_fit)
```

### Random Forest

```{r}
set.seed(1)
# Split data train and test
sample <- sample(c(TRUE, FALSE), nrow(data), replace=TRUE, prob=c(0.7,0.3))
train  <- data[sample, ]
test   <- data[!sample, ]

rf <- readRDS("rf_5050split.rds")
#rf <- randomForest(diabetes ~., data = train, importance = TRUE, 
#                   ntree = 300, do.trace = TRUE)
```

```{r}
set.seed(1)
yhat.rf = predict(rf, newdata = test)
# Confusion Matrix
(cm = confusionMatrix(test$diabetes, yhat.rf, positive = '1'))
# Variable Analysis
importance(rf)
varImpPlot(rf)
```

Accuracy: 0.7432

          Reference
Prediction    0    1
         0 7234 3432
         1 2006 8501
         
False negative: 0.1621
False positive: 0.0947
True negative: 0.3417
True positive: 0.4015

Sensitivity: 0.7124

BEST PERFORMING MODEL ???

If someone has diabetes, there's a 29% chance they are misclassified as not (3432 / (3432+8501)).



### Random Forest (removed GenHlth)

```{r}
data.subset <- subset(data, select = -c(GenHlth))

set.seed(1)
# Split data train and test
sample.subset <- sample(c(TRUE, FALSE), nrow(data.subset), replace=TRUE, prob=c(0.7,0.3))
train.subset  <- data.subset[sample.subset, ]
test.subset   <- data.subset[!sample.subset, ]

rf.subset <- randomForest(diabetes ~., data = train.subset, importance = TRUE, 
                          ntree = 300, do.trace = TRUE)
```

```{r}
set.seed(1)
yhat.rf.subset = predict(rf.subset, newdata = test.subset)
# Confusion Matrix
(cm.subset = confusionMatrix(test.subset$diabetes, yhat.rf.subset, positive = '1'))
# Variable Analysis
importance(rf.subset)
varImpPlot(rf.subset)
```
Accuracy: 0.7322

          Reference
Prediction    0    1
         0 7003 3612
         1 2099 8612

Sensitivity : 0.7045

Didn't perform as well as previous model, so ignore this.


#### Load Full Data

```{r}
data.large$diabetes <- as.factor(data.large$diabetes)
data.large$HighBP <- as.factor(data.large$HighBP)
data.large$CholCheck <- as.factor(data.large$CholCheck)
data.large$Smoker <- as.factor(data.large$Smoker)
data.large$Stroke <- as.factor(data.large$Stroke)
data.large$HeartDiseaseorAttack <- as.factor(data.large$HeartDiseaseorAttack)
data.large$PhysActivity <- as.factor(data.large$PhysActivity)
data.large$Fruits <- as.factor(data.large$Fruits)
data.large$Veggies <- as.factor(data.large$Veggies)
data.large$HvyAlcoholConsump <- as.factor(data.large$HvyAlcoholConsump)
data.large$AnyHealthcare <- as.factor(data.large$AnyHealthcare)
data.large$NoDocbcCost <- as.factor(data.large$NoDocbcCost)
data.large$GenHlth <- as.factor(data.large$GenHlth)
data.large$MentHlth <- as.factor(data.large$MentHlth)
data.large$DiffWalk <- as.factor(data.large$DiffWalk)
data.large$Sex <- as.factor(data.large$Sex)
data.large$Age <- as.factor(data.large$Age)
data.large$Education <- as.factor(data.large$Education)
data.large$Income <- as.factor(data.large$Income)
```

```{r}
set.seed(1)
# Remove duplicate rows (24206 duplicate)
data.large <- data.large[!duplicated(data.large), ]

# Split data into training and test
sample.large <- sample(c(TRUE, FALSE), nrow(data.large), replace=TRUE, prob=c(0.7,0.3))
train.large  <- data.large[sample.large, ]
test.large   <- data.large[!sample.large, ]
nrow(data.large)
nrow(train.large)
nrow(test.large)
```


### Random Forest on Full Dataset

```{r}
set.seed(1)
rf.large <- readRDS("rf_large_nosample.rds")
#rf.large <- randomForest(diabetes ~., data = train.large, importance = TRUE, 
#                   ntree = 300, do.trace = TRUE)
```

```{r}
#saveRDS(rf.large, "rf_large_nosample.rds")
```

```{r}
set.seed(1)
yhat.rf.large = predict(rf.large, newdata = test.large)
# Confusion Matrix
(cm.large = confusionMatrix(test.large$diabetes, yhat.rf.large, positive = '1'))
# Variable Analysis
importance(rf.large)
varImpPlot(rf.large)
```
Accuracy: 0.8518

          Reference
Prediction     0     1
         0 56887  1475
         1  8736  1787

False negative: 0.0214
False positive: 0.1268
True negative: 0.8258
True positive: 0.0259

Sensitivity: 0.5478


BEST PERFORMING MODEL ?

If someone has diabetes, there's a 45% chance they are misclassified as not (1475 / (1475+1787)).

### Random Forest on Full Data (Subset)

```{r}
# Removing least important variables based on Random Forest results
data.large.subset <- subset(data.large, select = -c(CholCheck))
```


```{r}
set.seed(1)
# Split data train and test
# sample.large.subset <- sample(c(TRUE, FALSE), nrow(data.large.subset), replace=TRUE, prob=c(0.7,0.3))
# train.large.subset  <- data.large.subset[sample.large.subset, ]
# test.large.subset   <- data.large.subset[!sample.large.subset, ]
# 
# rf.large.subset <- randomForest(diabetes ~., data = train.large.subset, importance = TRUE, 
#                                 ntree = 300, do.trace = TRUE)
```

```{r}
# set.seed(1)
# yhat.rf.large.subset = predict(rf.large.subset, newdata = test.large.subset)
# # Confusion Matrix
# (cm.large.subset = confusionMatrix(test.large.subset$diabetes, yhat.rf.large.subset))
# # Variable Analysis
# importance(rf.large.subset)
# varImpPlot(rf.large.subset)
```

Accuracy: 0.8519

          Reference
Prediction     0     1
         0 56872  1490
         1  8712  1811

False negative: 0.0216
False positive: 0.1264
True negative: 0.8256
True positive: 0.0263


Not really better than full model rf (rf.large), so ignore.


### Random Forest on Over-Sampled Data AFTER TRAINING

```{r}
set.seed(1)
# Sample from diabetes-1 classifications to balance dataset
class_0 = train.large[train.large['diabetes'] == 0,]
class_1 = train.large[train.large['diabetes'] == 1,]
class_1_over = class_1[sample(nrow(class_1), nrow(class_0), replace = TRUE), ]
train.large.sampled = rbind(class_1_over, class_0)
nrow(train.large.sampled)
```

```{r}
set.seed(1)
# rf.large.sampled <- randomForest(diabetes ~., data = train.large.sampled, importance = TRUE, 
#                                 ntree = 300, do.trace = TRUE)
```

```{r}
# set.seed(1)
# yhat.rf.large.sampled = predict(rf.large.sampled, newdata = test.large)
# # Confusion Matrix
# (cm.large.sampled = confusionMatrix(test.large$diabetes, yhat.rf.large.sampled, positive = '1'))
# # Variable Analysis
# importance(rf.large.sampled)
# varImpPlot(rf.large.sampled)
```
Accuracy: 0.7981

          Reference
Prediction     0     1
         0 49524  8838
         1  5071  5452

Sensitivity: 0.38153

REALLY BAD RESULTS-- DEFINITELY BAD MODEL


















